
*minilexer*: A Simple Lexer in R
==============================================================================


```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(minilexer)
})
knitr::opts_chunk$set(echo = TRUE)
```


`minilexer` provides some tools for simple tokenising/lexing and parsing text files.

Note: For complicated parsing (especially of computer programs) you'll probably want to use the more formally correct lexing/parsing provided by [the `rly` package on CRAN](https://cran.r-project.org/package=rly).

Installation
-----------------------------------------------------------------------------
```{r eval=FALSE}
devtools::install_bitbucket('coolbutuseless/minilexer')
```



Package Overview
-----------------------------------------------------------------------------

Current there is just one function provided: `minilexer::lex(text, patterns)`.

This function uses the user-defined regular expressions (`patterns`) to split 
`text` into a character vector of tokens.

The `patterns` argument is a named vector of character strings representing regular
expressions for elements to match within the text.  


ToDo
-----------------------------------------------------------------------------

* Add some parsing helpers
* Add vignettes for parsing some data formats, e.g.
    * chess games `.pgn`
    * scrabble games `.gcg`



Example: Lex a sentence into tokens
-----------------------------------------------------------------------------
```{r}
sentence_patterns <- c(
  word        = "\\w+", 
  whitespace  = "\\s+",
  fullstop    = "\\.",
  comma       = "\\,"
)

sentence = "Hello there, Rstats."

lex(sentence, sentence_patterns)
```




Example: Lex some simplified R code into tokens
-----------------------------------------------------------------------------

```{r}
R_patterns <- c(
  number      = "-?\\d*\\.?\\d+",
  name        = "\\w+",
  equals      = "==",
  assign      = "<-|=",
  plus        = "\\+",
  lbracket    = "\\(",
  rbracket    = "\\)",
  newline     = "\n",
  whitespace  = "\\s+"
)

R_code <- "x <- 3 + 4.2 + rnorm(1)"

lex(R_code, R_patterns)

```


