---
title: "Parsing 3d objects in OBJ format"
author: "mikefc"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(magrittr)
  library(dplyr)
  library(tidyr)
  library(glue)
  library(rgl)
  
  library(minilexer)
})

rgl::setupKnitr()

knitr::opts_chunk$set(echo = TRUE)
```






Example parser: `obj` format for 3d objects
------------------------------------------------------------------------------

A simple text file to store 3d objects is the Wavefront **obj** format.  The filetype
is well documented on the internet (e.g. 
[1](https://en.wikipedia.org/wiki/Wavefront_.obj_file), 
[2](http://paulbourke.net/dataformats/obj/), 
[3](https://www.cs.cmu.edu/~mbz/personal/graphics/obj.html)), and an example octahedron object
is show below which has 6 vertices and 8 faces.

```{r}
octahedron_obj <- '
# OBJ file created by ply_to_obj.c
#
g Object001

v  1  0  0
v  0  -1  0
v  -1  0  0
v  0  1  0
v  0  0  1
v  0  0  -1

f  2  1  5
f  3  2  5
f  4  3  5
f  1  4  5
f  1  2  6
f  2  3  6
f  3  4  6
f  4  1  6
'
```


The basic structure of a `.obj` file is:

* Comments start with `#` and continue to the end of the line
* There are symbols at the start of each line telling us what the data on the rest of the line represents, e.g.
    * `v` means this line defines a vertex and will be followed by 3 numbers representing the x, y, z coordinates.
    * `f` means this line defines a triangular face and the following 3 numbers indicate the 3 vertices which make up this face
    * `vn` means this line defines a vector for the direction of the normal at a vertex
* The format is more complicated than this, and I'm leaving out a lot of details, but this is enough to get the general idea.




Use `lex()` to turn the text into tokens
------------------------------------------------------------------------------

1. Start by defining the regular expression patterns for each element in the *obj* file.
2. Use `minilexer::lex()` to turn the *obj* text into tokens
3. Throw away whitespace, newlines and comments, since I'm not interested in them.

```{r}
obj_patterns <- c(
  comment    = '(#.*?)\n',  # assume comments take up the whole line
  number     = pattern_number,  # This regex is defined in `minilex` and matches most numeric values
  symbol     = '\\w+',
  newline    = '\n',
  whitespace = '\\s+'
)
```


Tokenising the `obj`
------------------------------------------------------------------------------

Split the `obj` text data into tokens, but then remove anything that we don't need 
to create the actual data structure representing the 3d object.

```{r}
tokens <- lex(octahedron_obj, obj_patterns)
tokens <- tokens[!(names(tokens) %in% c('whitespace', 'newline', 'comment'))]
tokens
```




Use a `TokenStream` to help turn the tokens into data
------------------------------------------------------------------------------

Initialise a `TokenStream` object to help us manipulate/interrogate the list of tokens we have.

```{r}
stream <- TokenStream$new(tokens)
```



Write a function to parse the lines which start with `f`
------------------------------------------------------------------------------

The lines which start with `f` encode a single triangular face. The numbers which
follow the `f` are the indicies of the vertices which make up the face.

To parse the lines which start with `f`: 

* make sure the current token is `f`
* keep consuming tokens as long as they are `numbers`
* when we run out of numbers, we consider this object parsed and return the data, in this case a numeric vector.


```{r}
parse_f <- function() {
  
  # make sure the current token is `f`
  stream$consume_token('symbol', 'f')
  
  # keep consuming tokens as long as they are `numbers`
  values <- stream$consume_tokens_of_type('number', c(3, 4))
  
  # when we run out of numbers, we consider this object parsed and 
  # return the data, in this case a numeric vector.
  as.numeric(values)
}
```



Write function to parse the lines which start with `g`, `v` and `vn`
------------------------------------------------------------------------------
```{r}
#-----------------------------------------------------------------------------
# Parse the 'group name' specification
#-----------------------------------------------------------------------------
parse_g <- function() {
  stream$consume_token('symbol', 'g')
  stream$consume_token()
}


#-----------------------------------------------------------------------------
# Parse the coordinates for a vertex. This may be 3 or 4 values, but 
# i'm just ignoring the 4th, and keeping the (x, y, z) 
#-----------------------------------------------------------------------------
parse_v <- function() {
  start_position <- stream$position
  stream$consume_token('symbol', 'v')
  values <- stream$consume_tokens_of_type('number', c(3, 4))
  if (length(values) == 4) {
    values <- head(values, -1)
  } 
  as.numeric(values)
}


#-----------------------------------------------------------------------------
# Parse the vector for that represents the normal at a vertex. 
# This may be 3 or 4 values, but i'm just ignoring the 4th, and keeping the (x, y, z) 
#-----------------------------------------------------------------------------
parse_vn <- function() {
  start_position <- stream$position
  stream$consume_token('symbol', 'vn')
  values <- stream$consume_tokens_of_type('number', c(3, 4))
  if (length(values) == 4) {
    values <- head(values, -1)
  } 
  as.numeric(values)
}
```


Write a top-level function containing a parse loop to keep extracting objects until we're done
------------------------------------------------------------------------------

* Check the current token
* Call the parser for that token
* Repeat

```{r}
parse_obj <- function() {
  obj   <- list()  # This is where we'll hold the parsed data.
  
  while (!is.na(stream$current_value())) {
    cv <- stream$current_value()
    if (cv == 'g') {
      parse_g()
    } else if (cv == 'v') {
      v <- parse_v()
      obj$v <- rbind(obj$v, v)
    } else if (cv == 'f') {
      f <- parse_f()
      obj$f <- rbind(obj$f, f)
    } else if (cv == 'vn') {
      vn <- parse_vn()
      obj$vn <- rbind(obj$vn, vn)
    } else {
      message <- glue("Parse error at position {stream$position}. Not understood: {stream$current_value()}")
      stop(message)
    }
  }
  
  obj
}

obj <- parse_obj()
```

The 3d object now exists as a list of data.frames (one for vertices and one for faces)

```{r}
obj
```



Post processing the data: Fortify/denormalise/tidy.
------------------------------------------------------------------------------

The text representation of the `obj` data is quite compact and avoids repetition
but this isn't quite in the right form for us to manipulate in R.

The following code turns this data into the `faces` data.frame which is slightly more
useful as each face has an actual ID, and the x, y and z co-ordinates of its 3 vertices (a, b, c)
are explicitly listed on each row i.e. we've created **a tidy data.frame** !

```{r}
#-----------------------------------------------------------------------------
# Fortify/denormalise/tidy the `f` and `v` data into `faces`
#-----------------------------------------------------------------------------
create_faces <- function(obj) {
  suppressWarnings({
    faces <- data.frame(obj$f) %>%
      set_names(c('a', 'b', 'c')) %>%
      mutate(face_id = seq(n())) %>%
      gather(idx, vert_id, -face_id) %>%
      arrange(face_id, idx) %>%
      as.tbl()
    
    verts <- data.frame(obj$v) %>%
      set_names(c('x', 'y', 'z')) %>%
      mutate(vert_id = seq(n())) %>%
      as.tbl()
  })
  
  faces %<>% left_join(verts, by='vert_id')
  
  faces
}

faces <- create_faces(obj)
```

```{r exho=FALSE}
faces %>% knitr::kable(caption='tidy faces data.structure')
```




Let's view the object!
------------------------------------------------------------------------------

Use your mouse to rotate and zoom the object.

```{r webgl=TRUE}
view3d(theta = 10, phi=15)
rgl::triangles3d(faces$x, faces$y, faces$z, col='grey')
```





