---
title: "Parsing 3d objects in OBJ format"
author: "mikefc"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(magrittr)
  library(dplyr)
  library(tidyr)
  library(glue)
  library(rgl)
  
  library(minilexer)
})

rgl::setupKnitr()

knitr::opts_chunk$set(echo = TRUE)
```






Example parser: `obj` format for 3d objects
------------------------------------------------------------------------------

A simple text file to store 3d objects is the Wavefront **obj** format.  The filetype
is well documented on the internet (e.g. 
[1](https://en.wikipedia.org/wiki/Wavefront_.obj_file), 
[2](http://paulbourke.net/dataformats/obj/), 
[3](https://www.cs.cmu.edu/~mbz/personal/graphics/obj.html)), and an example octahedron object
is show below which has 6 vertices and 8 faces.

```{r}
octahedron_obj <- '
# OBJ file created by ply_to_obj.c
#
g Object001

v  1  0  0
v  0  -1  0
v  -1  0  0
v  0  1  0
v  0  0  1
v  0  0  -1

f  2  1  5
f  3  2  5
f  4  3  5
f  1  4  5
f  1  2  6
f  2  3  6
f  3  4  6
f  4  1  6
'
```


The basic structure of a `.obj` file is:

* Comments start with `#` and continue to the end of the line
* There are symbols at the start of each line telling us what the data on the rest of the line represents, e.g.
    * `v` means this line defines a vertex and will be followed by 3 numbers representing the x, y, z coordinates.
    * `f` means this line defines a triangular face and the following 3 numbers indicate the 3 vertices which make up this face
    * `vn` means this line defines a vector for the direction of the normal at a vertex
* The format is more complicated than this, and I'm leaving out a lot of details, but this is enough to get the general idea.




Use `lex()` to turn the text into tokens
------------------------------------------------------------------------------

1. Start by defining the regular expression patterns for each element in the *obj* file.
2. Use `minilexer::lex()` to turn the *obj* text into tokens
3. Throw away whitespace, newlines and comments, since I'm not interested in them.

```{r}
obj_patterns <- c(
  comment    = '(#.*?)\n',  # assume comments take up the whole line
  number     = pattern_number,  # This regex is defined in `minilex` and matches most numeric values
  symbol     = '\\w+',
  newline    = '\n',
  whitespace = '\\s+'
)
```


Tokenising the `obj`
------------------------------------------------------------------------------

Split the `obj` text data into tokens, but then remove anything that we don't need 
to create the actual data structure representing the 3d object.

```{r}
tokens <- lex(octahedron_obj, obj_patterns)
tokens <- tokens[!(names(tokens) %in% c('whitespace', 'newline', 'comment'))]
tokens
```



