---
title: "Parsing scrabble games in gcg format"
author: "mikefc"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
suppressPackageStartupMessages({
  library(magrittr)
  library(dplyr)
  library(minilexer)
})
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Scrabble game format: gcg
------------------------------------------------------------------------------

The [gcg file format](http://www.poslfit.com/scrabble/gcg/) is a human readable 
representation of a Scrabble game.

In its most basic form, it has comments up the top (preceded by the `#` character) and then
the following rows represent a move by each player in turn.  

An example gcg file is show below:

```{r}
gcg_text <- '
#player1 Quackle Quackle Computer
#player2 David David Boys
#description Quackle Computer plays David Boys in Round 1 at the 2006 Human vs. Computer Showdown
#title 2006 Human vs. Computer Showdown Round 1
#incomplete
>Quackle: DEMJNOT  8d   JETON           +40   40
>David: ?EDYEIG   h2  rEDYEING        +64   64
>Quackle: BEDGMNP  7e   BEDIM           +26   66  BE, ET, DO
>David: HEALERS   j1  HEALERS         +75  139  BEDIMS
>Quackle: DFGINPS   k3  DIF             +29   95  AD, LI, EF
>David: COOAORS   l1  COOS            +28  167  ADO, LIS
>Quackle: EGNOPRS   m3  SPONGER         +92  187  ADOS, LISP
>David: AORWAVA  6c   AVOW            +37  204  OBE, WET
>Quackle: AEFMOVZ  8l   MEZE            +54  241
>David: AARTUNY   d8  JAUNTY          +32  236
>Quackle: ACFIOOV  1l   COOF            +27  268
>David: WALTIER  4c   WAILED          +20  256
>Quackle: AACEINV  3a   VIA             +22  290  AW
>David: IRUTRUT   a3  VIRTU            +9  265
>Quackle: AACEHLN  8a   EH              +42  332  VIRTUE
>David: QUBITUR  2b   BRUIT           +32  297  BI, RAW
>Quackle: AACILNR  9m   RAN             +16  348  ZA, EN
>David: PQUIEN? 13a   QUEY            +32  329
>Quackle: CALLIER   c13 EL               +2  350
>David: PINIR?N  1e   PIN             +11  340  PI, IT
>Quackle: ACEILOR 15a   CALORIE         +83  433  ELL
>David: TRAING? 14f   TRAdING         +67  407  TI, RE
>David:               (DATSXK)        +36  443
'
```

Use `lex()` to turn the text into tokens
------------------------------------------------------------------------------

1. Start by defining the regular expression patterns for each element in the gcg file.
2. Use `minilexer::lex()` to turn the *gcg* text into tokens

```{r}
gcg_patterns <- c(
  comment       = '(#.*?)\n',                 # Assume # only appears to denote comment to end of line
  newline       = '\n',
  whitespace    = '\\s+',
  player        = '>(.*?):',                  # start of each line with a `>`
  location      = '[a-o]\\d+|\\d+[a-o]|--|-', # Number first for horizontal words. -/-- for specials
  number        = minilexer::pattern_number,
  symbol        = '[-+\\w\\./\\?\\(\\)]+',
  comma         = ","
)

tokens <- minilexer::lex(gcg_text, gcg_patterns)
tokens <- tokens[!(names(tokens) %in% c('whitespace', 'newline', 'comment'))]
tokens[1:23]
```


Use a `TokenStream` to help turn the tokens into data
------------------------------------------------------------------------------

Initialise a `TokenStream` object to help us manipulate/interrogate the list of tokens we have.

```{r}
stream <- TokenStream$new(tokens)
```


Write a function to parse a single event line
------------------------------------------------------------------------------

Parsing a line which has a scrabble event on it consists of the following steps:

1. Check we are at the start of a line with a player's name.
2. The next symbol represents their rack of scrabble tiles.
3. The next symbol is a board location.
4. The next symbol is the word they played.
5. Don't bother extracting the scores - I'm not interested in them.
5. Return a single-row data.frame containing that information.


```{r}
parse_event <- function() {
  player_name <- stream$consume_token('player')
  rack        <- stream$consume_token('symbol')
  if (grepl("^\\(", rack)) {
    # If the rack is surrounded by () it is the
    # end-of-game rack for the other player.
    # Not handling this.
    return(NULL)
  } else {
    location <- stream$consume_token('location')
    play     <- stream$consume_token('symbol')
  }

  data_frame(player=player_name, rack=rack, location=location, play=play)
}
```


Write a parse loop to keep extracting events until we're done
------------------------------------------------------------------------------

1. For every line in the gcg file (after removing comments)
2. Call `parse_event()`
3. Stack all the events into a data.frame


```{r}
parse_gcg <- function() {
  events <- NULL
  
  while(!is.na(stream$current_value())) {
    if (stream$current_type() == 'player') {
      event <- parse_event()
      events <- rbind(events, event)
    } else {
      # silently skipping unhandled tokens
      stream$consume_token()
    }
  }
  
  events 
}

events <- parse_gcg()
```

```{r}
knitr::kable(events, caption="Scrabble game represented as a data.frame")
```


Manipulate the data
------------------------------------------------------------------------------

The whole point of parsing this data is to get it into a format that we can manipulate in R.

1. Interpret the `location` field as x and y coordinates
2. Add all the words to a game board
3. Display the board somehow

```{r}
#-----------------------------------------------------------------------------
# 1. Interpret the `location` field as x and y coordinates
#-----------------------------------------------------------------------------
events %<>% mutate(
  horizontal = grepl("^\\d", location),
  y          = as.integer(readr::parse_number(location)),
  xc         = stringr::str_extract(location, "[a-o]"),
  x          = charmatch(xc, letters)
)
```


```{r}
#-----------------------------------------------------------------------------
# 2. Add all the words to a game board (represented by a 15x15 character matrix)
#-----------------------------------------------------------------------------
board <- matrix('.', nrow=15, ncol=15)

add_word_to_board <- function(board, play, x, y, horizontal) {
  if (horizontal) {
    x <- seq(nchar(play)) + x - 1
  } else {
    y <- seq(nchar(play)) + y - 1
  }

  board[y, x] <- strsplit(play, '')[[1]]
  board
}

for (i in seq(nrow(events))) {
  event <- events[i,]
  board <- add_word_to_board(board, event$play, event$x, event$y, event$horizontal)
}
```


```{r}
#-----------------------------------------------------------------------------
# Print board
#-----------------------------------------------------------------------------
cat(paste(apply(board, 1, paste, collapse=' '), collapse="\n"))
```
